Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-23 report/report.tex
finished the paper

\usepackage{natbib}
\usepackage{multicol}
\bibliographystyle{plainnat}
% set path for bibliography

@@

% Paragraph for Motivating about Regionally Additive Models
Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models.
Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function (component) can be readily visualized and comprehended in isolation.
However, GAM's main limitation is that they cannot express interactions between features.
To mitigation this limitation, some approaches~\citep{lou2013accurate} extend them enabling pairwise interactions,
i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their x-by-design nature.
Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.

% Paragraph for Motivating about Regionally Additive Models
To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of x-by-design models,
that fits multiple GAMs to subregions of the feature space where interactions are minimized.
Our approach consists of a three-step pipeline.
First, we fit a black-box model to capture all high-order interactions.
Second, we identify the subregions where the black-box model is nearly locally additive.
Finally, we train a GAM specifically for each identified subregion.

\section{Background and motivation}
\label{sec:motivation}

Consider the black-box function \(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\)
with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
Although very simple, a GAM would fail to learn this mapping due to
the existence of the three-features interaction term $8x_2\when{x_1 > 0}\when{x_3=0}$.
As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $\hat{f}(\xb) \approx 2x_2$,
because in $\frac{1}{4}$ of the cases ($x_1 > 0 \text{ and } x_3 = 0$) the impact of $x_2$ to the output is $8x_2$,
and in the rest $\frac{3}{4}$ of the cases the impact of $x_2$ to the output is $0$.
However, if splitting the input space in two subregions we observe that \(f\) is additive in each one (regionally additive):
%
\begin{equation}
    \label{eq:regionally_additive}
    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 1 \\ 0 & \text{otherwise} \end{cases}
\end{equation}
%
Therefore, if we knew the appropriate subregions,
namely, \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}\)
and  \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}\),
we could split the impact of $x_2$ appropriately and fit the following model to the data:

@@

\section{The RAM framework}

The RAM framework consists of a three-step pipeline; (a) fit a black-box model (Section~\ref{subsec:fit_black_box}),
(b) identify subregions with minimal interactions (Section~\ref{subsec:regional_effect_methods}) and
(c) fit a GAM to each subregion (Section~\ref{subsec:fitting_gams}).
Throughout this section, we will use the following notation.
Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.
We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(/s = \{1, \ldots, D\} - s\) for the rest.
For convenience, we use \((x_s, \xcc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_{/s})\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).
%Finally, \(f^{\mathtt{<method>}}(x_s)\) denotes how \(\mathtt{<method>}\)
%defines the feature effect and \(\hat{f}^{\mathtt{<method>}}(x_s)\)
%how it estimates it from the training set.

@@

In this step, we use regional effect methods~\citep{herbinger2023decomposing, herbinger2022repid}
to identify the regions where the black-box function is (close to) regionally additive.
% Describe the goal of the regional effect methods
Regional effect methods yield for each individual feature \(s\), a set of \(T_s\) non-overlapping regions,
denoted as \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where \(\mathcal{R}_{st} \subseteq \Xcal_{/s}\).
Note that, the number of non-overlapping regions can be different for each feature ($T_s$),
the regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) are disjoint
and their union covers the entire feature space \(\Xcal_{/s}\).
The primary objective is to identify regions in which the impact of the \(s\)-th feature on the output is
\textit{relatively independent} of the values of the other features \(\xcc\).
To better grasp this objective, if we decompose the impact of the \(s\)-th feature on the output $y$ into two terms:
\(f_s(x_s, \xcc) = f_{s,ind}(x_s) + f_{s, int}(x_s, \xcc)\),
where \(f_{s,ind}(\cdot)\) represents the independent effect
and \(f_{s, int}(\cdot)\) represents the interaction effect,
the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) such that the interaction effect is minimized.
Regionally Additive Models (RAM) formulate the mapping \(\mathcal{X} \rightarrow \mathcal{Y}\) as:

\begin{equation}
\label{eq:ram_formulation2}
f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}) \quad \xb \in \Xcal^{\mathtt{RAM}}
\end{equation}
%
Equations~\ref{eq:ram_formulation} and~\ref{eq:ram_formulation2} are equivalent.
To gain a better understanding of the latter formulation, consider the toy example described in Section~\ref{sec:motivation}.
To minimize the impact of feature interactions, we need to divide feature \(x_2\) into two subregions,
\(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 1\}\) and \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 0\}\).
This division results in an augmented feature space \(\Xcal^{\mathtt{RAM}} = (x_1, x_{21}, x_{22}, x_3)\)
and a RAM formulation of the form: \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).

\subsubsection{Proposed Approach}

@@

\paragraph{DALE formulation}

DALE gets as input the black-box function \(f(\cdot)\)
and the dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
and returns the effect (impact) of a specific feature $s$ on the output $y$:
%
\begin{equation}  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \underbrace{\frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
    \mathcal{S}_k} \dfdx(\mathbf{x}^i)}_{\hat{\mu}(z_{k-1}, z_k)})
\end{equation}
%
For more details on the DALE method, please refer to the original paper~\citep{gkolemis2023dale}.
In the above equation, \(k_x\) is the index of the bin such that
\(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
is the set of the instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
In short, DALE computes the average effect (impact) of the feature \(x_s\) on the output,
by, first, dividing the feature space into $K$ equally-sized bins, i.e., \(z_0, \ldots, z_K\)
second, computing the average effect in each bin \(\hat{\mu}(z_{k-1}, z_k)\) (bin-effect) as the average of the instance-level effects inside the bin,
and, finally, aggregating the bin-level effects.

@@

In cases where there are strong interactions between the features,
the instance-level effects inside each bin deviate from the average bin-effect.
We can measure such deviation using the standard deviation of the instance-level effects inside each bin (bin-deviation):

\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}
%
and the interaction between the feature \(x_s\) and the rest of the features along the whole \(s\)-th dimension
with the aggregated bin-deviation:
\begin{equation}
  \label{eq:DALE_interaction}
  \mathcal{H}_s = \sqrt{ \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) }
\end{equation}
%
Eq.~\eqref{eq:DALE_interaction} measures the interaction between the feature \(x_s\) and the rest of the features.
It takes values in the range \([0, \infty)\) with zero indicating that \(x_s\) does not interact with the rest of the features,
i.e., the underlying black box function can be written as $f(\xb) = f_s(x_s) + f_{/s}(x_{/s})$.
In all other cases, $\mathcal{H}_s$ is greater than zero and the higher the value, the stronger the interaction.

A final detail, is that in order to have a more robust estimation of the bin-effect and the bin-deviation,
we use variable-size bins instead of the fixed-size ones in DALE.
In particular, we start with a dense fixed-size grid of bins and we iteratively merge the neighboring bins with similar
bin-effect and bin-deviation until all bins have at least $n_{\min}$ instances.
In this way, we can have a more accurate approximation of the bin-effect and the bin-deviation.

@@

In the same way that we can estimate the feature effect and the feature interactions for the $s$-th feature in the whole input space,
using Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction},
we can also estimate the effect and the interactions in a subregion of the input space \(\mathcal{R}_{st} \subset \mathcal{X}\).
We denote the equivalent regional qunatities as $f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$.
$f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ are defined exactly as in
Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction} respectively,
with the only difference that instead of using the whole dataset \(\mathcal{D}\) to compute the regional bin-effect $\hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k)$
and the regional bin-deviation $\hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)$,
we use only the instances that belong to the subregion \(\mathcal{R}_{st}\), i.e. $\xb^i: x_s^i \in \mathcal{S}_k \land x_{s}^i \in \mathcal{R}_{st}$.
Therefore, in order to minimize the interactions of a particular feature $s$ we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
that minimize:

\begin{equation}
  \label{eq:optimal_subregions}
  \begin{aligned}
    & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
    & & \mathcal{L} = \sum_{t=1}^{T_s} \mathcal{H}_{\mathcal{R}_{st}} \\
    & \text{subject to}
    & & \bigcup_{t=1}^{T} \mathcal{R}_{st} = \mathcal{X} \\
    & & & \mathcal{R}_{st} \cap \mathcal{R}_{s\tau} = \emptyset, \quad \forall t \neq \tau
  \end{aligned}
\end{equation}
%

@@

To minimize the objective of Eq.~\eqref{eq:SE}, we have developed a tree-based algorithm based on the approach proposed
by~\citep{herbinger2023decomposing}.
The core of the algorithm is outlined in Algorithm~\ref{alg:subregion_detection}.
We search for subregions independently for each feature $s \in \{1, \ldots, D\}$.
and we select the $T$ optimal splits, in a greedy manner.
To better comprehend the algorithm, let's consider the illustrative example of Section~\ref{sec:motivation}.
For feature $s=2$, the algorithm starts with the first level of the tree, with $/s = \{1, 3\}$ as candidate split-features.
Algorithm~\ref{alg:single_feature_subregion} identifies the optimal first-level split, with the following procedure.
Initially, it determines the candidate split points for each feature.

Since $x_1$ is a continuous feature, are generated as a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
Consequently, the candidate split points $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ define the candidate subregions
$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
As for $x_3$, being a categorical feature, the candidate split points are its unique values, i.e., $\{0, 1\}$,
and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.

For each candidate position, the algorithm splits the initial dataset into two subsets,
$[\mathcal{D}_1, \mathcal{D}_2]$ one for each subregion and computes the level of interaction
as the weighted sum of the interactions of the two subregions, as defined by Algorithm~\ref{alg:get_interaction}.
After the iteration over all points and all candidate positions,
the split point that minimizes the weighted level of interactions is selected.
In the illustrative example, the optimal first-level subregions are
$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
The algorithm then proceeds to the second level of the tree, where the only candidate feature is $x_3$.
In this step, the first split is considered fixed, i.e., the algorithm uses a list of two datasets,
and searches in an analogous manner for the optimal second-level split that will be applied on top of the first split,
leading to four subregions.
The algorithm proceeds in a similar manner, until it reaches the maximum depth $T$.

The output of algorithm~\ref{alg:subregion_detection} is a collection of $T$ splits for each feature $s$.
It is important to note that not all splits lead to a significant decrease in the level of interactions.
To address this, we post-process the splits by pruning any splits that do not result in a decrease of the
weighted interaction greater than a threshold value $\epsilon$ (set to $20\%$ drop in the experiments).
The pruned set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.

>>>>

\usepackage{natbib}
\usepackage{multicol}
\usepackage{algpseudocode}
\bibliographystyle{plainnat}
% set path for bibliography

@@

% Paragraph for Motivating about Regionally Additive Models
Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design
(x-by-design) models~\citep{rudin2019stop, ghassemi2021false}.
Their popularity stems from their inherent interpretability.
GAMs represent an aggregation of univariate functions, where the overall model can be expressed as
\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\).
Due to this structure, each individual univariate function (component) can be visualized and interpreted independently.
Consequently, understanding the behavior of the overall model simply requires visualizing all components, each with a one-dimensional plot.

However, GAMs have limited accuracy in cases where the outcome depends on multiple features simultaneously,
i.e., when the unknown predictive function includes terms that combine multiple features.
To mitigate this limitation, GA$^2$Ms~\citep{lou2013accurate} extend the traditional GAMs by adding pairwise interactions in their
    formulation, i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{s_1 \neq s_2} f_{s_1 s_2}(x_{s_1}, x_{s_2})\).
GA$^2$Ms are also x-by-design, because the user can visualize both the first-order ($1D$ plots) and
second-order (($2D$ plots)) components.
However, as the number of features increases, the number of second-order interactions grows exponentially,
making it impractical for users to remember and interpret a large number of two-dimensional plots.
Therefore, methods like GA$^2$Ms target on automatically selecting the most significant interaction terms.

Both GAMs and GA$^2$Ms have limitations in modeling interactions of more than two features,
and as far as we know, there is no existing work that focuses on x-by-design GAM-based models capable of
representing higher-order interactions.
The main reason for that is the difficulty of visualizing more than three terms simultaneously,
which would violate the x-by-design principle.

To address this limitation, we propose a new class of x-by-design models called Regionally Additive Models (RAMs).
Since in the general case, it is infeasible to visualize terms with more than two variables
RAMs focus on learning terms with the following conditional structure:
$f(x_{s_1} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for first-degree interactions and
$f(x_{s_1}, x_{s_2} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for second-degree interactions.
The symbol $\when{x_{c_1}}$ denotes the condition that the feature $x_{c_1}$ takes a specific value or belongs to a specific range.

To better grasp the idea, consider a prediction task where the outcome depends, among others,
on a combination $f(x_1, x_2, x_3)$ of three features:
$x_1 \in [20, 80]$ (age), $x_2 \in [0, 40]$ (years in work), and $x_3 \in \{True, False\}$ (married).
Both GAM and GA$^2$M would fail to accurately learn this term of the underlying predictive function.
However, the three-feature effect can be decomposed in two sets of second-degree conditional terms based
on the marital status: $f_{1}(x_1, x_2 | x_3 = True)$ and $f_{2}(x_1, x_2 | x_3 = False)$.
In this way, RAM can accurately represent $f$ through learning two second-degree conditional terms, one for each marital status.
Furthermore, the two sets of terms can be visualized and interpreted as using two-dimensional plots.
It is worth noting that the conditional terms can also include numerical features.
For example, it could be more accurate to learn instead a set of four first-degree terms, conditioned on the marital status and the years in work:
$f_{1}(x_1 | x_2 < 10, x_3 = True)$,
$f_{2}(x_1 | x_2 \geq 10, x_3 = True)$,
$f_{3}(x_1 | x_2 < 10, x_3 = False)$, and
$f_{4}(x_1 | x_2 \geq 10, x_3 = False)$, which can be visualized and interpreted as four one-dimensional plots.

To adhere to the x-by-design principle, RAMs should be able to automatically identify the
most significant conditional terms.
As the number of these terms increases, it becomes difficult for users to retain and interpret numerous plots
associated with each feature or pair of features.
Therefore, RAMs use Regional Effect Plots~\citep{herbinger2023decomposing} to identify a small set
of conditional terms that have the greatest impact in minimizing feature interactions.
The RAM framework consists of three key steps.
First, a black-box model is fitted to capture all high-order interactions.
Then, the subregions where the black-box model exhibits near-local additivity are identified using Regional Effect Plots.
Finally, a GAM component is fit to each identified subregion.

The main contributions of this paper are as follows:

\begin{itemize}
    \item We formulate a new class of x-by-design models called Regionally Additive Models (RAMs).
    \item We propose a generic framework for learning RAMs and we propose a novel method for identifying the most significant conditional terms.
    \item We demonstrate the effectiveness of RAMs in modeling high-order interactions on a synthetic toy example and two real-world datasets.
\end{itemize}

\section{Motivation}
\label{sec:motivation}

Consider the black-box function \(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\)
with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
Although very simple, GAM and GA$^2$M would fail to learn this mapping due to
the the three-features interaction term.
As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $\hat{f}(\xb) \approx 2x_2$,
because in $\frac{1}{4}$ of the cases ($x_1 > 0 \text{ and } x_3 = 0$) the impact of $x_2$ to the output is $8x_2$,
and in the rest $\frac{3}{4}$ of the cases the impact of $x_2$ to the output is $0$.
However, if splitting the input space in two subregions we observe that \(f\) is additive in each one (regionally additive):
%
\begin{equation}
    \label{eq:regionally_additive}
    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 1 \\ 0 & \text{otherwise} \end{cases}
\end{equation}
%
Therefore, if we knew the appropriate subregions,
namely, \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}\)
and  \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}\),
we could split the impact of $x_2$ appropriately and fit the following model to the data:

@@

\section{RAM formulation}
\label{sec:ram_formulation}

\paragraph{Notation.}
Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.
We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(/s = \{1, \ldots, D\} - s\) for the rest.
For convenience, we use \((x_s, \xcc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_{/s})\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).
%Finally, \(f^{\mathtt{<method>}}(x_s)\) denotes how \(\mathtt{<method>}\)
%defines the feature effect and \(\hat{f}^{\mathtt{<method>}}(x_s)\)
%how it estimates it from the training set.

The RAM consists of a three-step pipeline; (a) fit a black-box model (Section~\ref{subsec:fit_black_box}),
(b) identify subregions with minimal interactions (Section~\ref{subsec:regional_effect_methods}) and
(c) fit a GAM component to each subregion (Section~\ref{subsec:fitting_gams}).

In step (b), we use regional effect methods~\citep{herbinger2023decomposing, herbinger2022repid}
to identify the regions where the black-box function is (close to) regionally additive.
% Describe the goal of the regional effect methods
Regional effect methods yield for each individual feature \(s\), a set of \(T_s\) non-overlapping regions,
denoted as \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where \(\mathcal{R}_{st} \subseteq \Xcal_{/s}\).
Note that, the number of non-overlapping regions can be different for each feature ($T_s$),
the regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) are disjoint
and their union covers the entire feature space \(\Xcal_{/s}\).
The primary objective is to identify regions in which the impact of the \(s\)-th feature on the output is
\textit{relatively independent} of the values of the other features \(\xcc\).
To better grasp this objective, if we decompose the impact of the \(s\)-th feature on the output $y$ into two terms:
\(f_s(x_s, \xcc) = f_{s,ind}(x_s) + f_{s, int}(x_s, \xcc)\),
where \(f_{s,ind}(\cdot)\) represents the independent effect
and \(f_{s, int}(\cdot)\) represents the interaction effect,
the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) such that the interaction effect is minimized.
Regionally Additive Models (RAM) formulate the mapping \(\mathcal{X} \rightarrow \mathcal{Y}\) as:

\begin{equation}
\label{eq:ram_formulation2}
f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}) \quad \xb \in \Xcal^{\mathtt{RAM}}
\end{equation}
%
Equations~\ref{eq:ram_formulation} and~\ref{eq:ram_formulation2} are equivalent.
To better understand of the formulations, consider the toy example described in Section~\ref{sec:motivation}.
To minimize the impact of feature interactions, we need to divide feature \(x_2\) into two subregions,
\(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 1\}\) and \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 0\}\).
Using Eq.~\ref{eq:ram_formulation}, RAM formulation is:
\(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_2) \when{x_1 > 0 \text{ and } x_3 = 1} + f_{22}(x_2) \when{x_1 \leq 0 \text{ or } x_3 = 0} + f_3(x_3)\).
Using Eq.~\ref{eq:ram_feature_space}, we should first define the augmented feature space
\(\Xcal^{\mathtt{RAM}} = (x_1, x_{21}, x_{22}, x_3)\),
where \(x_{21} = x_2 \when{x_1 > 0 \text{ and } x_3 = 1}\) and \(x_{22} = x_2 \when{x_1 \leq 0 \text{ or } x_3 = 0}\)
and then RAM formulation is: \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).

\section{RAM framework}
\label{sec:ram_framework}

@@

@@

\paragraph{DALE}

DALE gets as input the black-box function \(f(\cdot)\)
and the dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
and returns the effect (impact) of the $s$-th feature $s$ on the output $y$:
%
\begin{equation}  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \underbrace{\frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
    \mathcal{S}_k} \dfdx(\mathbf{x}^i)}_{\hat{\mu}(z_{k-1}, z_k)})
\end{equation}
%
For more details on the DALE method, please refer to the original paper~\citep{gkolemis2023dale}.
In the above equation, \(k_x\) is the index of the bin such that
\(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
is the set of the instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
In short, DALE computes the average effect (impact) of the feature \(x_s\) on the output,
by, first, dividing the feature space into $K$ equally-sized bins, i.e., \(z_0, \ldots, z_K\)
second, computing the average effect in each bin \(\hat{\mu}(z_{k-1}, z_k)\) (bin-effect) as the average of the instance-level effects inside the bin,
and, finally, aggregating the bin-level effects.

@@

In cases where there are strong interactions between the features,
the instance-level effects inside each bin deviate from the average bin-effect (bin-deviation).
We can measure such deviation using the standard deviation of the instance-level effects inside each bin:

\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}
%
The bin-deviation is a measure of the interaction between the feature \(x_s\) and the rest of the features inside the \(k\)-th bin.
Therefore, we can measure the global interaction between the feature \(x_s\) and the rest of the features along the whole \(s\)-th dimension
with the aggregated bin-deviation:
\begin{equation}
  \label{eq:DALE_interaction}
  \mathcal{H}_s = \sqrt{ \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) }
\end{equation}
%
Eq.~\eqref{eq:DALE_interaction} outputs values in the range \([0, \infty)\) with zero indicating that \(x_s\) does not interact with any other feature,
i.e., the underlying black box function can be written as $f(\xb) = f_s(x_s) + f_{/s}(x_{/s})$.
In all other cases, $\mathcal{H}_s$ is greater than zero and the higher the value, the stronger the interaction.

A final detail, is that in order to have a more robust estimation of the bin-effect and the bin-deviation,
we use variable-size bins instead of the fixed-size ones in DALE.
In particular, we start with a dense fixed-size grid of bins and we iteratively merge the neighboring bins with similar
bin-effect and bin-deviation until all bins have at least a minimum number of instances.
In this way, we can have a more accurate approximation of the bin-effect and the bin-deviation.

@@

In the same way that we can estimate the feature effect (Eq.~\eqref{eq:DALE_accumulated_mean_est}) and
the feature interactions (Eq.~\eqref{eq:DALE_interaction}) for the $s$-th feature in the whole input space,
we can also estimate the effect and the interactions in a subregion of the input space \(\mathcal{R}_{st} \subset \mathcal{X}\).
We denote the equivalent regional qunatities as $\hat{f}^{\mathtt{DALE}}_{st}(x_s)$ and $\mathcal{H}_{st}$.
$f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ are defined exactly as in
Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction} respectively,
with the only difference that instead of using the whole dataset \(\mathcal{D}\), to compute the regional bin-effect $\hat{\mu}_{st}(z_{k-1}, z_k)$
and the regional bin-deviation $\hat{\sigma}_{st}^2(z_{k-1}, z_k)$,
we use $\mathcal{D}_{st}$ which includes only the instances that belong to the subregion \(\mathcal{R}_{st}\),
i.e., $\mathcal{D}_{st} = \{\xb^i: x_s^i \in \mathcal{S}_k \land \xc^i \in \mathcal{R}_{st}\}$.
Therefore, in order to minimize the interactions of a particular feature $s$ we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
that minimizes the following objective:

\begin{equation}
  \label{eq:optimal_subregions}
  \begin{aligned}
    & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
    & & \mathcal{L}_s = \sum_{t=1}^{T_s} \frac{| \mathcal{D}_{st} |}{|\mathcal{D}|} \mathcal{H}_{st} \\
    & \text{subject to}
    & & \bigcup_{t=1}^{T} \mathcal{R}_{st} = \mathcal{X} \\
    & & & \mathcal{R}_{st} \cap \mathcal{R}_{s\tau} = \emptyset, \quad \forall t \neq \tau
  \end{aligned}
\end{equation}
%
In Eq.~\eqref{eq:optimal_subregions}, the objective function is the weighted sum of the regional interactions $\mathcal{H}_{st}$,
where the weights are the number of instances in each subregion.
In this way, we give more importance to the subregions that contain more instances.
The first constraint ensures that the subregions cover the whole input space and the second constraint ensures that the subregions are disjoint.

@@

The core of the method is outlined in Algorithm~\ref{alg:ram}.
First, we fit a differentiable black box model to the data (Step 1) and we copute the Jacobian matrix w.r.t. the input features (Step 2).
Then we search for a set of subregions by minimizing the objective of Eq.~\eqref{eq:optimal_subregions}
for each feature $s$ independently (Steps 3-4-5).
Based on the optimal subregions, we define the extended feature space (Step 6) and we fit a GAM in the extended feature space (Step 7).

For solving Eq.~\eqref{eq:optimal_subregions}, we have developed a tree-based algorithm
based on the approach proposed by~\citep{herbinger2023decomposing}, which we describe in detail in Algorithm~\ref{alg:subregion_detection}.
To describe the algorithm, we define some additional notation: $\mathcal{R}_s^l$ is the set of optimal subregion of the $s$-th feature at level $l$ of the tree.
Since at each level of the tree we divide the input space into two subregions, at level $l$ we have $2^l$ subregions,
i.e., $\mathcal{R}_s^l = \{\mathcal{R}_{st}\}_{t=1}^{2^l}$.
Equivalently, $\mathcal{L}_s^l$ is the optimal objective value of Eq.~\eqref{eq:optimal_subregions} at level $l$ of the tree.
Although the algorithm can search for an abritrary number of subregions per feature,
in order to preserve the smooth interpretation of the method,
we limit the maximum depth of the tree to $L=3$ levels,
which stands for a maximum of $T = 2^L = 8$ subregions per feature.
In general, the user can control the trade-off between the interpretability and the accuracy of the method by changing the maximum depth of the tree.
Note that with three splits, we already have an interaction term of
four ($f(x_s | \when{x_{c_1}}, \when{x_{c_2}}, \when{x_{c_3}}$) or
five ($f(x_{s_1}, x_{s_{2}} | \when{x_{c_1}}, \when{x_{c_2}}, \when{x_{c_3}}$) features.

To describe how the algorithm finds the optimal splits at each level $l$, let's consider the illustrative example of Section~\ref{sec:motivation}.
For feature $s=2$, the algorithm starts with $/s = \{1, 3\}$ as candidate split-features for the first level of the tree.
For each candidate split-feature, the algorithm determines the candidate split positions.
Since $x_1$ is a continuous feature, the candidate splits postions are a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
where $P$ is a hyperparameter of the algorithm, set to $10$ in the experiments.
Therefore, the candidate positions are $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ each on defining two subregions,
$\mathcal{R}_{21} = \{ (x_1, x_3) : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ (x_1, x_3) : x_1 > p \}$.
As for $x_3$, being a categorical feature, the candidate split points are its unique values, i.e., $\{0, 1\}$,
and the corresponding subregions are $\mathcal{R}_{21} = \{ (x_1, x_3) : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ (x_1, x_3) : x_3 \neq 0 \}$.
Each candidate position, creates a corresponding dataset $[\mathcal{D}_{21}, \mathcal{D}_{22}]$,
and the algorithm computes the weighted level of interactions $\mathcal{H}_{21}$ and $\mathcal{H}_{22}$ for each dataset.
After iterating over all features and all candidate positions for each feature,
it selects the split point that minimizes the weighted level of interactions.
In the illustrative example, the optimal first-level split is based on $x_3$ and the optimal split point is $p = 0$.
The algorithm next proceeds to the second level, where the only candidate feature is $x_3$.
In this step, the first split is considered fixed so the optimal second split is applied to the subregions $\mathcal{R}_{21}$ and $\mathcal{R}_{22}$,
creating four subregions in total.
The algorithm continues in a similar manner, until it reaches the maximum depth $T$ or the drop in the weighted level of interactions is below a threshold $\epsilon$ (set to $20\%$ drop in the experiments).
